Do NBA Teams Avoid Trading within Their Own Division?

### 2023-05-31 Manuscript.Rmd
Add Appendix with model specification details

@@

$$P( Y = y ) = \frac{\exp (
\sum_{s=2}^S \theta_s g_1 (y, s) +
\sum_{d=1}^S \theta_d g_1 (y, d) +
\theta_h g_2 (y))}{\kappa(\theta_s, \theta_d, \theta_h)}$$

where the sufficient statistics are:

- $g_1(y, s)$ -- number of trade participations in season $s$
- $g_1(y, d)$ -- number of trade participations by teams in division $d$
- $g_2(y)$ -- number of trades involving teams from the same divisions

The model is constrained to within-season dyads only.

@@

$$P( Y = y ) = \frac{\exp (
\text{CMP} 
\sum_{s=2}^S \theta_s g_1 (y, s) +
\theta_h g_2 (y))}{\kappa(\theta_s, \theta_d, \theta_h)}$$

where the sufficient statistics are:

-

The model is constrained to within-season dyads only.

>>>>

## Details of model specification {.appendix}

Basic notation:

- $N$ -- number of unique team-season pairs
- $i, j \in \{1, ..., N\}$ -- team playing in a season (a team-season pair)
- $\{s_1, ..., s_i, ..., s_N\}$ -- season of each team-season pair
- $s_i \in \{1, ..., s, ..., S\}$ -- seasons
- $\{d_1, ..., d_i, ..., d_N\}$ -- division of each team-season pair
- $d_i \in \{1, ..., d, ..., D\}$ -- divisions
- $[y_{ij}]_{N \times N}$ -- A block-diagonal adjacency matrix of trades between teams in  seasons. For simplicity, but abusing notation slightly, $y_{ij}$ is the number of trades in a season between teams $i$ and $j$ in case of valued ERGMs and, in the case of binary ERGM, trade count dichotomised to $\{0,1\}$ (no trade vs any number of trades).

In words, each node of the network corresponds to a team playing in the particular season. Nodes are grouped into seasonal blocks such that trades can happen only within seasons (ties within blocks) but cannot happen between seasons (no ties between blocks).

The models use the following sufficient statistics:

- *Season-specific propensity to trade* represented by the number of trade participations by teams in season $s$:

$$g_1(y, s) = \sum_{i,j:\; s_i=s \,\vee\, s_j=s} y_{ij}$$

- *Division-specific propensity to trade* represented by the number of trade participations by teams in division $d$:

$$g_1(y, d) = \sum_{i,j:\; d_i = d \,\vee\, d_j=d} y_{ij}$$

- *Propensity towards within-division trading* represented by the number of trades involving teams from the same division:

$$g_2(y) = \sum_{i,j:\; d_i = d_j} y_{ij}$$

- *Over-/under-dispersion* statistic which modifies the reference measure into a Conway-Maxwell-Poisson-reference ERGM [c.f. @krivit2012, sec. 5.2.3]:

$$g_3(y) = \sum_{i,j} \log(y_{ij}!)$$

Pooled binary ERGM is specified as:

$$P( Y = y ) = \frac{\exp \left(
\sum_{s=2}^S \theta_s g_1 (y, s) +
\sum_{d=1}^D \theta_d g_1 (y, d) +
\theta_2 g_2 (y) \right)}{\kappa}$$

Pooled valued ERGM  is specified as:

$$P( Y = y ) = \frac{\exp \left(
\theta_3 g_3 (y) + 
\sum_{s=1}^S \theta_s g_1 (y, s) +
\theta_2 g_2 (y) \right)}{\kappa}$$

@@

@@
